Fuels, fire weather, and Rx fire behavior in North Dakota

### 2021-02-12 writeups/FuelsWeatherFirebehavior.Rmd
update with new data from KY

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
FilePath = "C:/Users/devan/GitHubProjects/SpatialFireBehavior"
load(paste0(FilePath, '/data/AnalysisData.Rdata'))
load(paste0(FilePath, '/r objects/response_CIs.Rdata'))
load(paste0(FilePath, '/r objects/imp_raw.Rdata'))
source(paste0(FilePath, '/r script/CustomThemes.R'))
source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
```

@@

```{r data_prep, eval=FALSE}
 AllData <-  
    read_csv(paste0(FilePath, "data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, MaxWindSpeed, 
               LAI, FMC, KgHa)

@@

# Combine, filter into final tibble for analysis 
  AnalysisData  <- 
    full_join(DataMeans, ROS) %>%
              left_join(SoilTemp) %>%
      filter( ros <= 40, 
              MaxC >= 40) %>%
      rename(FuelMoisture = FMC, 
             SoilMaxC = SoilC) %>%
        mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
                                      FuelMoisture, NA), 
               FuelMoisture = FuelMoisture * 100) %>%
      separate(FireCode, into = c("location", "block", "pasture", 
                           "patch", "year"), 
               remove = F)
```

```{r fig.width = 10, fig.height = 8, fig.cap="A scatterplot matrix of all raw data used in this analysis."}
# Scatterplot matrix  
  GGally::ggpairs(AnalysisData, 
                  columns = 9:18, 
                  upper = 'blank')
```

@@

Therefore I used the multiple imputation method in the **mice** package to fill in missing values. 
The method creates 100 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
These results are pooled into a single set of results.

@@

```{r ADMB_mods, eval=FALSE}
# Max soil surface temperature 
  # Fit model
    soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
                            FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=filter(imp_d, SoilMaxC <= 3.0),
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10) )
  # Retrieve regression results
    soil_terms <-   
      confint(soil_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(soil_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

# Maximum aboveground temp (plant canopy)  
  # Fit model
    canopy_ADMB <- glmmADMB::glmmadmb(MaxC ~ RH + MaxWindSpeed +
                                        FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_d,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results
  canopy_terms <-   
    confint(canopy_ADMB) %>%
    as.data.frame %>%
    rownames_to_column("term") %>%
    rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
    full_join(
      coef(canopy_ADMB) %>%
        as.data.frame() %>% 
        rownames_to_column("term") %>%
        rename(estimate = '.') )

# Rate of spread 
  # Fit model 
    ros_ADMB <- glmmADMB::glmmadmb(ros ~ RH + MaxWindSpeed +
                                    FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_d,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results  
    ros_terms <-   
      confint(ros_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(ros_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

@@

**Rate of spread** significantly increased with windspeed ($z = 2.61, P = 0.01$) and decreased with greater fuel moisture content ($z = -3.92, P < 0.001$). 
**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -5.48, P < 0.001$) and increased with fuel load ($z = 5.16, P < 0.001$) and wind speed ($z = 2.10, P = 0.04$).
**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.77, P < 0.01$), which likely indicates higher intensity fires that burn down through litter. 
There was a marginally-significant trend for higher fuel moisture to reduce temperature at the soil surface, as well ($z = -1.85, P = 0.06$).

Further work should include quantifying total heat exposure via infrared sensors.

@@

No difference in fire behavior patterns between Hettinger and Central Grasslands.

```{r}
# Testing fire weather against PCA
  envfit(fb_pca ~ MaxWindSpeed + AirTemp + dpC + RH + VPD, 
         data = imp_raw, 
         choices = c(1:3), 
         strata = imp_raw$location)
```

RH and VPD had a statistically-significant associations with variation in fire behavior, but even these relationships were weak.

```{r pca_dat}  
  pca_v <- envfit(fb_pca ~ RH + VPD, imp_raw, 
              choices = c(1:3), 
              strata = imp_raw$location)

  # get ord data for ggplotting
  pca_gg <- gg_ordiplot(fb_pca, groups = imp_raw$location, 
                        plot=FALSE)

@@

  # Vector 
  pca_vd <- scores(pca_v,  "vectors") %>%
            as.data.frame %>% 
            round(3) %>%
            as_tibble(rownames="gradient")

@@

```{r pca_gg, fig.height=4, fig.width=6, fig.cap='Principal Components Analysis of fire behavior data (response variables in red) for prescribed burns on rangeland at Hettinger (H) and Central Grasslands  (CG) Research Extension Centers. No difference between locations and only the relative humidty (RH) gradient had a significant association with variation in fire behavior.' }
# ggplot PCA
  ggplot() + theme_bw(16) + 
    labs(x="PC 1", 
         y="PC 2") + 
    geom_vline(xintercept = 0, lty=3, color="darkgrey") +
    geom_hline(yintercept = 0, lty=3, color="darkgrey") +
    geom_point(data=pca_scores$spiders, 
                        aes(x=PC1, y=PC2, 
                            shape=Group, 
                            colour=Group), 
                        size=2)  +   
    geom_segment(data=pca_scores$spiders, 
                 aes(x=cntr.x, y=cntr.y,
                     xend=PC1, yend=PC2, 
                     color=Group), 
                 size=1.2, 
                 alpha = 0.75) +
    geom_point(data=pca_scores$spiders, 
               aes(x=PC1, y=PC2, fill=Group), 
               colour="grey40", 
               pch=21, 
               size=3, 
               stroke=2, alpha = 0.75) +
    geom_segment(data=pca_scores$vectors, 
                 aes(x=0, y=0, 
                     xend=PC1*1.6, 
                     yend=PC2*1.6),
                 arrow=arrow(length = unit(0.03, "npc")),
                 lwd=1.5) +
  geom_label(data=pca_scores$spiders,
               aes(x=cntr.x,
                   y=cntr.y,
                   label=Group,
                   color=Group),
               fontface="bold", size=4,
               label.size = 0,
               label.r = unit(0.5, "lines")) +
    geom_text(data=pca_scores$vectors,
              aes(x=PC1*1.75, 
                  y=PC2*1.75, 
                  label=gradient), 
              nudge_x = 0.06, 
              nudge_y =-0.05, 
              size=6, fontface="bold") + 
    geom_label(data=pca_scores$species, 
                aes(x=PC1/3, 
                    y=PC2/3, 
                    label=response), 
                label.padding=unit(0.25,"lines"),
                label.size = 0.5, 
                fontface="bold", 
               color="darkred") +
    scale_color_manual(values = cbPal5[1:2]) + 
    scale_fill_manual(values = cbPal5[1:2])  +
    theme(panel.grid=element_blank(), 
          legend.position="none")

@@

>>>>

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
FilePath = "C:/Users/devan/GitHubProjects/SpatialFireBehavior"
AnalysisData <- readr::read_csv(paste0(FilePath, "/data/AnalysisDataKY.csv"))
load(paste0(FilePath, '/r objects/response_CIs.Rdata'))
load(paste0(FilePath, '/r objects/imp_raw.Rdata'))
load(paste0(FilePath, '/r objects/pca_ev.Rdata'))
source(paste0(FilePath, '/r script/CustomThemes.R'))
source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
```

@@

```{r data_prep, eval=FALSE}
###
# Note: Data prep steps included for reference. 
# Data used in analysis, AnalysisDataKY.csv, has been modified outside of
# this script to update missing data from another spreadsheet. 
 AllData <-  
    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, MaxWindSpeed, 
               LAI, FMC, KgHa)

@@

# Combine, filter into final tibble for analysis 
  AnalysisData  <- 
    full_join(DataMeans, ROS) %>%
              left_join(SoilTemp) %>%
      filter( ros <= 40, 
              MaxC >= 40) %>%
      rename(FuelMoisture = FMC, 
             SoilMaxC = SoilC) %>%
        mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
                                      FuelMoisture, NA), 
               FuelMoisture = FuelMoisture * 100) %>%
      separate(FireCode, into = c("location", "block", "pasture", 
                           "patch", "year"), 
               remove = F)
###
```

```{r fig.width = 10, fig.height = 8, fig.cap="A scatterplot matrix of all raw data used in this analysis."}
# Scatterplot matrix  
AnalysisData %>%
  select(-LAI) %>%
  GGally::ggpairs(columns = 9:18, 
                  upper = 'blank')
```

@@

Therefore I used the multiple imputation method in the **mice** package to fill in missing values. 
The method creates 500 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
These results are pooled into a single set of results.

```{r imp, eval=FALSE}
# Calculate imputed datasets
  imp <- AnalysisData %>% 
          select(-LAI) %>%
              mutate(across(plot:array, ~as.character(.))) %>%
               mice(m=500, seed = 23109, print=F)

# Return completed datset   
  imp_raw <- 
    complete(imp) %>%
            as_tibble() %>% 
              mutate(across(location:array, as.factor))

# Scale response variables for regression analysis   
  imp_sc <- 
    imp_raw %>%
     mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F)))
```

@@

```{r ADMB_mods, eval=FALSE}
# Max soil surface temperature 
  # Fit model
    soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
                            FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=filter(imp_sc, SoilMaxC <= 3.9),
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10) )
  # Retrieve regression results
    soil_terms <-   
      confint(soil_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(soil_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

# Maximum aboveground temp (plant canopy)  
  # Fit model
    canopy_ADMB <- glmmADMB::glmmadmb(MaxC ~ RH + MaxWindSpeed +
                                        FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_sc,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results
  canopy_terms <-   
    confint(canopy_ADMB) %>%
    as.data.frame %>%
    rownames_to_column("term") %>%
    rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
    full_join(
      coef(canopy_ADMB) %>%
        as.data.frame() %>% 
        rownames_to_column("term") %>%
        rename(estimate = '.') )

# Rate of spread 
  # Fit model 
    ros_ADMB <- glmmADMB::glmmadmb(ros ~ RH + MaxWindSpeed +
                                    FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_sc,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results  
    ros_terms <-   
      confint(ros_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(ros_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

@@

**Rate of spread** significantly increased with windspeed ($z = 3.24, P < 0.01$) and decreased with greater fuel moisture content ($z = -5.84, P < 0.001$). 
**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -6.45, P < 0.001$) and increased with fuel load ($z = 5.49, P < 0.001$) and wind speed ($z = 2.82, P < 0.01$).
**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.52, P = 0.01$) and fuel moisture ($z = -3.22, P < 0.01$), which likely indicates higher-intensity fires that burn down through litter.

@@

No difference in fire behavior patterns between Hettinger and Central Grasslands; this is expected.

```{r envfit, eval=FALSE}
# Testing fire weather against PCA
 envfit(fb_pca ~ MaxWindSpeed + AirTemp + dpC + RH + VPD, 
         data = imp_raw, 
         choices = c(1:3), 
         strata = imp_raw$location)
```

```{r echo=FALSE}
pca_ev
```

RH and VPD had statistically-significant associations with variation in fire behavior, but even these relationships were weak.

```{r pca_dat}  
  # get ord data for ggplotting
  pca_gg <- gg_ordiplot(fb_pca, groups = imp_raw$location, 
                        plot=FALSE)

@@

  # Vector 
  pca_vd <- scores(pca_ev,  "vectors") %>%
                as.data.frame %>% 
                rownames_to_column("gradient") %>%
              filter(gradient %in% c("RH", "VPD")) %>%
              select(-PC3) %>%
            as_tibble( )

@@

```{r pca_gg, fig.height=4, fig.width=6, fig.cap='Principal Components Analysis of fire behavior data (response variables in red) for prescribed burns on rangeland at Hettinger (H) and Central Grasslands  (CG) Research Extension Centers. No difference between locations and only the vapor pressure deficit (VPD) and relative humidty (RH) gradients had significant associations with variation in fire behavior.' }
# ggplot PCA
  ggplot() + theme_bw(16) + 
    labs(x="PC 1", 
         y="PC 2") + 
    geom_vline(xintercept = 0, lty=3, color="darkgrey") +
    geom_hline(yintercept = 0, lty=3, color="darkgrey") +
    geom_point(data=pca_scores$spiders, 
                        aes(x=PC1, y=PC2, 
                            shape=Group, 
                            colour=Group), 
                        size=2)  +   
    geom_segment(data=pca_scores$spiders, 
                 aes(x=cntr.x, y=cntr.y,
                     xend=PC1, yend=PC2, 
                     color=Group), 
                 size=1.2, 
                 alpha = 0.75) +
    geom_point(data=pca_scores$spiders, 
               aes(x=PC1, y=PC2, fill=Group), 
               colour="grey40", 
               pch=21, 
               size=3, 
               stroke=2, alpha = 0.75) +
    geom_segment(data=pca_scores$vectors, 
                 aes(x=0, y=0, 
                     xend=PC1*1.6, 
                     yend=PC2*1.6),
                 arrow=arrow(length = unit(0.03, "npc")),
                 lwd=1.5) +
  geom_label(data=pca_scores$spiders,
               aes(x=cntr.x,
                   y=cntr.y,
                   label=Group,
                   color=Group),
               fontface="bold", size=4,
               label.size = 0,
               label.r = unit(0.5, "lines")) +
    geom_text(data=pca_scores$vectors,
              aes(x=PC1*1.75, 
                  y=PC2*1.75, 
                  label=gradient), 
              nudge_x = 0.06, 
              nudge_y =-0.05, 
              size=6, fontface="bold") + 
    geom_label(data=pca_scores$species, 
                aes(x=PC1/3, 
                    y=PC2/3, 
                    label=response), 
                label.padding=unit(0.25,"lines"),
                label.size = 0.5, 
                fontface="bold", 
               color="darkred") +
    scale_color_manual(values = cbPal5[1:2]) + 
    scale_fill_manual(values = cbPal5[1:2])  +
    theme(panel.grid=element_blank(), 
          legend.position="none")

@@

It is interesting that the gradient in temperature at the soil surface (SoilMaxC) is orthogonal to both canopy temperature and rate of spread (MaxC and ros, respectively). 
This confirms trends observed in the regression analysis that the conditions most conducive to fast, high-intensity fires are not the same that result in surface heating.

Managers might be able to use high windspeeds to mitigate soil heating when fuels are dry, or focus on conditions with low RH to ensure litter burns down to the soil surface.
